---
title: "Confirmatory testing: matching claims to evidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confirmatory testing: matching claims to evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 150, fig.width = 7,
                      fig.height = 5.6, out.width = "100%",
                      fig.align = "center")
set.seed(2026)
library(lagdynamics)
options(digits = 3)
has <- function(p) requireNamespace(p, quietly = TRUE)
```

A fitted model is an estimate, not a finding. The Dynalytics framework
(Saqr, Lopez-Pernas, and Misiejuk, 2026) formalises this as a *scientific
contract*: every analytical claim must be matched by evidence appropriate
to its structure, scope, and complexity. `lagdynamics` provides the
confirmatory testing battery that discharges that contract for
lag-sequential models, where each edge is a tested departure from
independence.

The battery pairs a kind of claim with a kind of evidence:

| Claim | Evidence | Function |
|---|---|---|
| a *specific transition* is real / how precise it is | edge-level uncertainty | `certainty_lsa()`, `bootstrap_lsa()` |
| a *significant transition* is not fragile | robustness to information loss | `stability_lsa()` |
| the *whole network* is reproducible | structural reliability | `reliability_lsa()` |
| the structure is *more than chance* | an assumption-free null | `permute_lsa()` |
| two *groups* differ | inference under exchangeability | `compare_lsa()`, `bayes_compare_lsa()` |

We use the bundled `engagement` data (138 students, weekly engagement
states) for the single-network tests, and a long event log with a real
group for the comparison.

```{r fit}
fit <- lsa(engagement)
transitions(fit, significant = TRUE)
```

The adjusted residual already tests each transition against the
independence model, but that test rests on large-sample assumptions that
sequential data can violate. The battery supplies evidence that does not
rely on them.

# A specific transition: edge-level uncertainty

A claim about one edge requires an estimate of how much that edge could
vary. `bootstrap_lsa()` resamples whole sequences and re-fits;
`certainty_lsa()` derives the same uncertainty analytically from a
Dirichlet-Multinomial posterior. The two agree when the population is
homogeneous; the bootstrap is preferred for a mixture, because resampling
sequences preserves within-sequence dependence that the analytic model
treats as independent.

```{r edge}
as.data.frame(bootstrap_lsa(fit, R = 200)) |> head(4)   # resampling CIs
as.data.frame(certainty_lsa(fit)) |> head(4)            # analytic CIs
```

The bootstrap forest shows every edge's interval at once; tightly pinned
edges support stronger claims.

```{r forest, eval = has("ggplot2"), fig.height = 6.5}
plot(bootstrap_lsa(fit, R = 200))
```

# A significant transition: robustness to information loss

`stability_lsa()` repeatedly drops a fraction of the cases and re-fits,
recording how often each significant edge stays significant. A high
stability supports reading the edge as a property of the process; a low
one marks it as sample-dependent.

```{r stability}
as.data.frame(stability_lsa(fit, R = 200)) |> head(4)
```

# The whole network: structural reliability

`reliability_lsa()` raises the claim from a single edge to the entire
network. It repeatedly splits the sequences into two halves, fits a model
on each, and correlates the edge-weight vectors. A high average
correlation indicates the network is reproducible within the sample.

```{r reliability}
reliability_lsa(fit, R = 50)
```

# More than chance: an assumption-free null

`permute_lsa()` shuffles the event order to build the null of no
sequential structure, giving a p-value that does not depend on the
adjusted residual's large-sample approximation.

```{r permute}
as.data.frame(permute_lsa(fit, R = 200)) |> head(4)
```

# Two groups: inference under exchangeability

Comparing groups is the claim that most invites over-interpretation:
two networks estimated separately almost always look different. The
question is whether the difference exceeds what arises by chance when the
group labels carry no information.

`tna::group_regulation_long` is a long event log with a recorded
achievement group; the same `actor` / `action` / `time` grammar fits it in
one call.

```{r group, eval = has("tna")}
log <- tna::group_regulation_long
gfit <- lsa(log, actor = "Actor", action = "Action", time = "Time",
            group = "Achiever")
```

`compare_lsa()` answers it by permutation under exchangeability: it
shuffles the labels to build a reference distribution for the per-edge
difference and a single omnibus statistic.

```{r compare, eval = has("tna")}
cmp <- compare_lsa(gfit, R = 500, adjust = "BH")
cmp
as.data.frame(cmp) |> subset(significant) |> head(4)
```

```{r barrel, eval = has("tna") && has("ggplot2"), fig.height = 6.5}
plot(cmp)
```

`bayes_compare_lsa()` is the Bayesian counterpart: instead of a
significance decision it reports, for each edge, the posterior mean
difference and a credible interval, so the evidence is a plausible range
rather than a verdict.

```{r bayes, eval = has("tna")}
bayes_compare_lsa(gfit, seed = 1)
```

# In short

The contract is one rule applied at every scope: match the claim to the
evidence. A descriptive reading of an edge needs only the fit; a stronger
claim needs the test that targets exactly its structure.

```r
certainty_lsa(fit); bootstrap_lsa(fit)   # a specific edge
stability_lsa(fit)                       # a significant edge under case-dropping
reliability_lsa(fit)                     # the whole network
permute_lsa(fit)                         # more than chance
compare_lsa(gfit); bayes_compare_lsa(gfit)  # a group difference
```
