---
title: "Interoperability with tna and Nestimate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interoperability with tna and Nestimate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 150, fig.width = 7,
                      fig.height = 5.6, out.width = "100%",
                      fig.align = "center")
set.seed(2026)
library(lagseq)
options(digits = 3)
has <- function(p) requireNamespace(p, quietly = TRUE)
```

`lagseq` belongs to the Dynalytics ecosystem alongside `tna` (transition
network analysis), Nestimate (network estimation), `cograph` (rendering),
and `TraMineR`. The ecosystem separates estimation from rendering and
analysis, and the packages share a common data grammar and object
schemas. In practice this means two things: `lsa()` can read sequences
straight out of those packages' objects, and a fitted `lsa` model
converts back into them in one call.

# Into lagseq: fit from another package's object

`lsa()` recognises sequence-bearing objects from `tna`
(`tna`, `tna_data`, `tna_seq_data`), Nestimate, and `TraMineR`
(`stslist`). It pulls the sequences out and fits, so no manual extraction
is needed.

## A tna object

`tna` and `lsa()` share the same long-format grammar
(`actor` / `action` / `time` / `order` / `session`). Prepare the data with
`tna`, build a model, and hand either object to `lsa()`.

```{r tna-in, eval = has("tna")}
log <- tna::group_regulation_long
prepared  <- tna::prepare_data(log, actor = "Actor",
                               action = "Action", time = "Time")
tna_model <- tna::tna(prepared)

lsa(prepared)     # fit from the prepared sequence object
lsa(tna_model)    # or from a built tna model -- same sequences, same fit
```

## A TraMineR state sequence object

A `TraMineR` `stslist` (the standard state-sequence object) is read
directly.

```{r tramineR-in, eval = has("TraMineR")}
sts <- TraMineR::seqdef(engagement)
lsa(sts)
```

# Out of lagseq: hand a fit to another toolkit

A fitted model is a directed weighted network, so it converts to the
native object of either network package. Estimation stays in `lagseq`;
the downstream analysis runs in the toolkit built for it.

```{r fit}
fit <- lsa(engagement)
```

## To a tna network

`lsa_to_tna()` returns a `tna` object. Choose the edge weight with
`weights` (`"prob"` for a transition-probability network, `"count"` for a
frequency one). From there, tna's analysis verbs apply.

```{r to-tna, eval = has("tna")}
tn <- lsa_to_tna(fit, weights = "prob")
tna::centralities(tn) |> head()
```

## To an igraph

`as.igraph()` returns an `igraph`, so the full igraph toolbox is
available.

```{r to-igraph, eval = has("igraph")}
ig <- igraph::as.igraph(fit, weights = "prob")
igraph::betweenness(ig)
```

# Shared rendering: the transition network is a tna model

Because estimation and rendering are separate layers, the transition
network view of an `lsa` fit *is* a `tna` model: `lsa_to_tna()` builds it
on the fly and tna's own plot method renders it (coloured nodes,
initial-probability arcs, weighted directed edges).

```{r render, eval = has("tna")}
plot(fit, type = "network", weights = "prob")
```

The residual network, by contrast, is rendered by `cograph` and keeps the
lag-sequential meaning (each edge a tested departure from independence).
One fit, two rendering layers, one shared object model.

# One grammar across the ecosystem

The practical payoff is a single mental model. The arguments that sequence
a raw log are the same in `lsa()` and in `tna::prepare_data()`:

```r
tna::prepare_data(log, actor =, action =, time =, order =, session =)
lsa(log,               actor =, action =, time =, order =, session =)
```

So an analysis can move between transition-network tooling and
lag-sequential testing without reshaping the data, and an object built in
one package is valid input to the other.
