---
title: "Lag transition networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lag transition networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 150, fig.width = 7,
                      fig.height = 5.4, out.width = "100%",
                      fig.align = "center")
set.seed(2026)
library(lagdynamics)
options(digits = 3)
has <- function(p) requireNamespace(p, quietly = TRUE)
```

# Lag-sequential analysis

Lag-sequential analysis studies the order in which categorical events occur.
Given a series of events, the method examines pairs that follow one another
and tests whether a given event is followed by another event more or less
often than the order of events alone would predict. The event types become
the states of the process, and the transitions between them become the
object of analysis. A lag transition network is the representation of these
tested transitions: the states are nodes, a directed edge from one state to
another represents the transition between them, and the weight on the edge
measures how far the observed transition departs from what independence would
predict.

The method rests on a comparison between two quantities for every ordered
pair of states. The observed count is the number of times the first state is
immediately followed by the second. The expected count is the number that
would occur if the next state did not depend on the current one, and it is
built from the base rates of the two states alone. The adjusted residual
expresses the difference between the two on a standardised scale. A positive
residual identifies a transition that occurs more often than expected, which
marks a behavioural regularity above the base rates of its states. A negative
residual identifies a transition that occurs less often than expected, which
marks an avoidance. A residual near zero identifies a transition that occurs
about as often as chance predicts. This tutorial develops these quantities,
fits a model to a real data set, and interprets the network that results.

# The data and the fitted model

The `ai_long` data set is a long-format event log of coded AI-side actions in
human--AI coding sessions, with one row per event.

```{r data}
data(ai_long)
head(ai_long)
```

Four columns enter the analysis. The `code` column holds the action, the
event type whose transitions are modelled. The `order_in_session` column
orders events within a session. The `session_id` column identifies one
uninterrupted sequence, and the `project` column groups the sessions.
Transitions are counted within a session only, so the last event of one
session and the first of the next are never treated as a pair; the `session`
argument enforces this boundary.

The `lsa()` function fits the model, taking the four columns as named
arguments.

```{r fit}
fit <- lsa(ai_long,
           actor   = "project",
           session = "session_id",
           action  = "code",
           order   = "order_in_session")
fit
```

The printed model reports the number of states, transitions, and sequences,
the omnibus test of independence, the strongest over-represented
transitions, and the initial-state distribution. The `ai_long` process has 8
states and 8123 within-session transitions drawn from 428 sessions, so the
model estimates a compact 8-by-8 transition system from a large number of
observed moves; the residuals it reports are therefore repeated tendencies
in the process rather than isolated episodes.

# Reading the model

The fitted model exposes its contents through accessor functions, each of
which returns a table.

The initial-state distribution records the share of sessions that begin in
each state.

```{r initial}
initial(fit)
```

The initial-state distribution places most of its mass on `Investigate`,
whose value of 0.715 means that about 71% of sessions begin with an
investigation action. `Delegate` begins 18% of sessions and `Execute` 6%,
while `Repair` and `Report` never begin a session. The process almost always
opens by gathering information.

Node activity records how often each state sends and receives transitions.

```{r nodes}
nodes(fit)
```

`Execute` sends 3090 transitions and receives 3233, more than any other
state, which makes it the most frequently visited state in the process.
`Ask`, `Report`, and `Explain` are visited rarely. The magnitude of these
totals is exactly what the expected-count calculation controls for, so that a
busy state is not credited with meaningful transitions merely for being busy.

The omnibus test evaluates the whole table against independence and is the
first result to consult.

```{r tests}
tests(fit)
```

The likelihood-ratio statistic is $G^2 = 2168$ on 49 degrees of freedom,
with a p-value below machine precision. The order of actions is therefore not
random: the next action depends on the current one. A non-significant
omnibus test would remove any basis for interpreting individual transitions,
so it is checked before them.

The `transitions()` function returns one row per directed edge. Its
`direction` argument selects over- or under-represented transitions and its
`sort` argument orders them by strength.

```{r over}
transitions(fit, direction = "over", sort = "strength")
```

`Investigate -> Plan` is the strongest over-represented transition. It
occurred 977 times, its transition probability is 0.44, and its adjusted
residual is 33.4. The probability means that planning follows investigation
44% of the time, and the lift of 2.21 means this is 2.2 times more often than
independence predicts; investigation is regularly converted into planning.
`Ask -> Explain` carries a lift of 7.51, so asking is followed by explaining
about seven times more often than chance. `Execute -> Execute` has a
probability of 0.51, so execution is followed by execution about half the
time and tends to occur in runs. `Delegate -> Plan` and `Explain -> Report`
describe related regularities.

```{r under}
transitions(fit, direction = "under", sort = "strength")
```

`Plan -> Plan` is the strongest under-represented transition, with an
adjusted residual of -17.9 and a lift of 0.16. Planning is therefore very
rarely repeated immediately: a planning action is followed by another
planning action only about one-sixth as often as expected. The result does
not mean planning is uncommon; it means planning is normally followed by a
different action. `Investigate -> Execute` (residual -16.6) is also
under-represented, so investigation seldom leads straight to execution. Read
together with the over-represented `Investigate -> Plan`, these avoidances
show a separation between gathering information and acting on it, with a
planning step in between.

# The network

The network represents each state as a node and each transition as a
directed edge, with edge width proportional to the weight and an edge that
loops to its own node marking a self-transition. Three networks can be drawn
from the same model, differing in the quantity they place on the edges. The
two lag-sequential networks come first: the residual network, whose edges are
signed adjusted residuals (blue over-represented, red avoided), and the
Yule's Q network, which shows the same signed departures on a fixed $[-1, 1]$
scale that does not grow with sample size. The probability network, which is
the Transition Network Analysis view, comes last.

The residual network draws the tested departures from independence. The full
network draws every edge:

```{r net-residual-full, eval = has("cograph")}
plot_transitions(fit, weights = "residuals", decimals = 1)
```

```{r net-residual-full-skip, echo = FALSE, eval = !has("cograph")}
message("Install 'cograph' to draw the residual network.")
```

Because a residual grows with sample size and the data set has many
transitions, most edges are significant and the full network is dense. The
`top` argument keeps the strongest edges by absolute residual:

```{r net-residual-top, eval = has("cograph")}
plot_transitions(fit, weights = "residuals", top = 12, decimals = 1)
```

The pruned residual network has a directed shape that the edge weights make
explicit. `Investigate -> Plan` is the heaviest edge (residual 33.4), and its
reverse `Plan -> Investigate` is also over-represented (residual 5.6);
investigation and planning therefore form a mutual pair. `Plan -> Execute`
then runs forward (residual 6.3), while its reverse `Execute -> Plan` is red
(residual -14.5), so the link from planning to execution is one-directional.
`Execute -> Execute` is a heavy self-loop (residual 16.1), so the process
remains in execution once it arrives. The red edges `Plan -> Plan` and
`Investigate -> Execute` confirm that planning does not repeat and that
investigation does not jump straight to execution.

The Yule's Q network shows the same signed departures on a bounded scale, so
its edge values are comparable across datasets and across groups of different
sizes. The full network again draws every edge, and `top` keeps the
strongest:

```{r net-yulesq-full, eval = has("cograph")}
plot_transitions(fit, weights = "yules_q", decimals = 2)
```

```{r net-yulesq-full-skip, echo = FALSE, eval = !has("cograph")}
message("Install 'cograph' to draw the Yule's Q network.")
```

```{r net-yulesq-top, eval = has("cograph")}
plot_transitions(fit, weights = "yules_q", top = 12, decimals = 2)
```

The Yule's Q network preserves the shape of the residual network, the same
over-represented and avoided edges, but its weights are bounded association
values rather than test statistics, which is the form to use when the
networks of two groups are to be compared.

The probability network draws the same transitions weighted by
$P(\text{to} \mid \text{from})$, the share of moves out of a state that reach
each destination. This is the network of Transition Network Analysis (TNA); a
thin ring around each node shows the initial probability of starting in that
state.

```{r net-prob, eval = has("cograph")}
plot_transitions(fit, weights = "prob")
```

```{r net-prob-skip, echo = FALSE, eval = !has("cograph")}
message("Install 'cograph' to draw the probability network.")
```

The probability network foregrounds the most frequent transitions:
`Delegate -> Plan` (0.62), `Execute -> Execute` (0.51), `Ask -> Explain`
(0.48), `Plan -> Execute` (0.47), and `Investigate -> Plan` (0.44). It is not
equivalent to the lag-sequential networks. A transition probability lies
between 0 and 1, records how often a transition occurs, and reaches its
smallest value at zero, so it cannot represent a transition that occurs less
often than expected. A signed residual or Yule's Q can. The under-represented
edges of the residual network, such as `Execute -> Plan`, have no counterpart
in the probability network. The probability network describes where the
process goes; the lag-sequential networks identify which transitions depart
from chance, and in which direction.

# Inference

The adjusted residual is an analytic test, and its conclusions can be checked
against the data by resampling. The bootstrap resamples whole sequences with
replacement, refits the model on each resample, and records a confidence
interval for every edge together with whether the edge keeps its sign.

```{r bootstrap}
boot <- bootstrap_lsa(fit, R = 1000)
boot
```

The bootstrap finds 43 of the 64 possible edges to be stable, meaning their
sign is preserved across the resamples. A stable edge can be reported with
confidence, whereas an unstable edge might not reappear in a new sample even
when its residual is large.

Analytic certainty offers a closed-form alternative to the bootstrap. The
`certainty_lsa()` function models each state's outgoing transitions as
Dirichlet--Multinomial, which yields a posterior distribution, a credible
interval, and a certainty decision for each transition probability without
resampling.

```{r certainty}
cert <- certainty_lsa(fit)
cert
```

Analytic certainty classifies 26 of 64 edges as certain, meaning their
transition probabilities are estimated precisely. The bootstrap and the
analytic estimate agree on the overall picture: the frequent transitions are
estimated with confidence, and the rare transitions are not.

# Summary

The analysis of `ai_long` establishes that the order of actions is not random
($G^2 = 2168$, $p < 0.001$) and that the process has a definite shape.
Sessions almost always begin in `Investigate` (initial probability 0.71).
`Investigate` and `Plan` form a mutual pair, feed forward into `Execute`
through the one-directional `Plan -> Execute`, and terminate in `Execute`,
which is the busiest state (in-strength 2.30), occurs in runs
(`Execute -> Execute`, probability 0.51), and rarely returns to planning.
Planning does not repeat (`Plan -> Plan`, residual -17.9) and investigation
seldom leads straight to execution (`Investigate -> Execute`, residual
-16.6), so a planning step separates information gathering from acting on it.
Of the 64 possible transitions, 40 are significant, 43 are sign-stable under
the bootstrap, and 26 are certain under the Bayesian estimate, so a
transition merits reporting when it is significant, sizeable, and
reproducible. The residual network and the probability network remain
complementary throughout: the probability network describes where the
process tends to go, and the residual network identifies which transitions
occur more or less often than chance.

# References

Bakeman, R., & Gottman, J. M. (1997). *Observing interaction: An
introduction to sequential analysis* (2nd ed.). Cambridge University Press.

Haberman, S. J. (1973). The analysis of residuals in cross-classified
tables. *Biometrics*, 29(1), 205--220.

Saqr, M., López-Pernas, S., & Tikka, S. (2025). Mapping relational dynamics
with transition network analysis: A primer and tutorial. In *Advanced
Learning Analytics Methods and Tutorials*.
