---
title: "Get started with lagseq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with lagseq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dpi = 150,
                      fig.width = 7, fig.height = 5.6, out.width = "100%",
                      fig.align = "center")
set.seed(1)
library(lagseq)
options(digits = 3)
```

`lagseq` answers one question about a stream of categorical events:
**given that one state just happened, does another follow more — or
less — often than chance?**

Reading a fit is predictable: each result is `verb(fit)`, and each
returns a tidy one-row-per-observation `data.frame`. We use the built-in
`group_regulation` data — 2,000 collaborative-learning sessions coded
into nine regulation actions.

## Fit

```{r fit}
fit <- lsa(group_regulation)
fit
```

## Reading a fit — one verb per result

```{r transitions}
transitions(fit) |> head()   # one row per from -> to transition
```

```{r nodes-tests-initial}
nodes(fit)                  # one row per state, with in/out totals
tests(fit)                  # tablewise independence tests (G^2, chi^2)
initial(fit)                # initial-state distribution
```

```{r summary}
summary(fit)                # one-row overview (prints rich view, returns tidy)
```

### Picking transitions

The transitions you want are arguments to `transitions()`, not separate
functions:

```{r filters}
transitions(fit, significant = TRUE) |> head()    # significant only
transitions(fit, direction = "over")  |> head(4)  # over-represented
transitions(fit, direction = "under") |> head(4)  # avoided
transitions(fit, min_count = 500)     |> head(4)  # frequently observed
```

## Plotting

Every view comes from one verb, `plot(fit, type = )`: `"heatmap"`
(default), `"network"`, `"chord"`, `"sunburst"`, and `plot()` on a
`bootstrap_lsa()` result for the bootstrap forest. Grouped fits plot one
panel per group. The default is the residual heatmap.

```{r plot-residuals, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(fit)                     # residual heatmap (default)
```

See `vignette("plotting", package = "lagseq")` for the full gallery.

## Getting data in

`lsa()` takes many shapes; supply `actor`/`action`/`time` to sequence a
raw long event log on the fly.

```{r input}
lsa(c("a", "b", "a", "c", "b"))               # one sequence (vector)
lsa(list(c("a", "b", "a"), c("b", "c", "a")))  # several sequences

log <- data.frame(
  user = c("u1", "u1", "u1", "u2", "u2"),
  act  = c("login", "read", "quiz", "login", "read"),
  ts   = as.POSIXct(c("2025-01-01 10:00", "2025-01-01 10:01",
                      "2025-01-01 10:02", "2025-01-02 09:00",
                      "2025-01-02 09:01"), tz = "UTC"),
  stringsAsFactors = FALSE)
lsa(log, actor = "user", action = "act", time = "ts")
```

The canonical input form and the raw transition table are also
available on their own:

```{r data-helpers}
lsa_data(c("a", "b", "a", "c"))               # canonical sequence object
lsa_transitions(c("a", "b", "a", "c"))        # raw transition counts
```

## Lags

`lsa()` takes any integer lag (negative = predecessors, `0` = self).
`lag_profile()` tracks one transition across lags; `lsa_lags()` fits a
whole profile.

```{r lags}
lag_profile(group_regulation, "plan", "consensus", lags = 1:3)
as.data.frame(lsa_lags(group_regulation, lags = 1:2)) |> head(3)
```

## Groups

Give `lsa()` one label per sequence; every verb then returns grouped,
tidy results with a leading `group` column.

```{r groups}
g <- rep(c("A", "B"), length.out = nrow(group_regulation))
gfit <- lsa(group_regulation, group = g)
transitions(gfit, significant = TRUE) |> head(4)
nodes(gfit) |> head(4)      # every verb works grouped: leading `group` column
summary(gfit)               # one tidy row per group
```

## Inference

Resampling quantifies uncertainty. Each returns an object whose
`as.data.frame()` is tidy.

```{r inference}
as.data.frame(bootstrap_lsa(fit, R = 100)) |> head(3)   # edge CIs
reliability_lsa(fit, R = 30)                            # split-half reliability (summary)
as.data.frame(reliability_lsa(fit, R = 30)) |> head(3)  # per-replicate draws (tidy)
```

```{r inference-more, eval = FALSE}
permute_lsa(fit, R = 100)     # permutation p-values
stability_lsa(fit, R = 100)   # case-drop stability
```

## Engines

The classical engine is the default; others target specific designs,
and you can register your own.

```{r engines}
list_lsa_engines()
lsa(group_regulation, engine = "classical")$method
```

Convenience wrappers call each engine directly: `lsa_classical()`,
`lsa_two_cell()`, `lsa_bidirectional()`, `lsa_parallel_dominance()`,
`lsa_nonparallel_dominance()`. Register a custom one with
`register_lsa_engine()` (and remove it with `unregister_lsa_engine()`).

## Structural zeros

Forbid impossible cells with a 0/1 matrix; forbidden cells become `NA`,
never a false "expected".

```{r structural-zeros}
K <- nrow(fit$obs)
fz <- lsa(group_regulation, structural_zeros = 1 - diag(K))
transitions(fz, significant = TRUE) |>
  subset(from == to)    # self-transitions: none, they are NA
```

## To a network toolkit

A fit is a directed weighted network, so it converts in one call to the
native object of a network package (these need the package installed).

```{r tna, eval = requireNamespace("tna", quietly = TRUE)}
tna::centralities(lsa_to_tna(fit, weights = "prob")) |> head()
```

```{r igraph, eval = requireNamespace("igraph", quietly = TRUE)}
igraph::betweenness(igraph::as.igraph(fit, weights = "prob"))
```

## In short

```r
fit <- lsa(group_regulation)          # fit
transitions(fit)                      # tidy edge table
transitions(fit, significant = TRUE)  # the transitions that matter
nodes(fit); tests(fit); initial(fit)  # other results, same shape
summary(fit)                          # one-row overview
plot(fit)                             # residual heatmap
lag_profile(group_regulation, "plan", "consensus")  # lag profile
lsa(group_regulation, group = g)      # grouped
bootstrap_lsa(fit)                    # uncertainty
```

See `?lsa` for every input form and `?lsa_to_tna` for the network bridge.
