---
title: "Plotting LSA fits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting LSA fits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dpi = 100,
                      fig.width = 6.5, fig.height = 5.2, out.width = "100%",
                      fig.align = "center")
set.seed(1)
library(lagseq)
options(digits = 3)
has_gg <- requireNamespace("ggplot2", quietly = TRUE)
has_cg <- requireNamespace("cograph", quietly = TRUE)
fit <- lsa(group_regulation)                       # single fit
gf  <- lsa(group_regulation,                       # grouped fit
           group = rep(c("A", "B"), length.out = nrow(group_regulation)))
```

`lagseq` has one plotting verb, **`plot(fit, type = )`**. Every view
shares a single palette: **warm (red) = over-represented, cool (blue) =
avoided**, the adjusted residual against chance. The views differ only in
geometry, never in meaning.

| `type` | view | backend |
|---|---|---|
| `"heatmap"` (default) | `from x to` grid | ggplot2 |
| `"network"` | transition network | cograph `splot()` |
| `"chord"` | chord diagram | cograph `plot_chord()` |
| `"sunburst"` | polar rose | ggplot2 |
| — | bootstrap CI forest | ggplot2 (`plot()` on a bootstrap) |

**Single or grouped.** `plot()` works the same on one fit and on a
grouped fit: a grouped fit draws **one full-size figure per group** by
default (`combined = FALSE`); pass `combined = TRUE` to tile them into a
single compact figure. Each view below is shown both ways. The ggplot
views need `ggplot2`; the network and chord need `cograph` (both
*Suggests*).

## Sunburst — `type = "sunburst"`

A polar portrait of the transition table. Source states are named along
the inner ring. Two styles: **`"rose"`** (default; equal slots, bar
height = frequency, fill = residual) and **`"wedge"`** (angular width =
frequency share, tiny wedges omitted).

```{r sunburst, eval = has_gg}
plot(fit, type = "sunburst")                       # single fit, rose
plot(fit, type = "sunburst", style = "wedge")      # frequency wedges
```

```{r sunburst-group, eval = has_gg}
plot(gf, type = "sunburst")                        # grouped: one per group
```

## Heatmap — `type = "heatmap"` (default)

Rows are source states ordered by outgoing volume; the salient cell of
each row is bold. `which` picks the matrix.

```{r heatmap, eval = has_gg}
plot(fit)                     # single fit, adjusted residuals (default)
plot(fit, which = "prob")     # transition probabilities
```

```{r heatmap-group, eval = has_gg}
plot(gf)                      # grouped: one heatmap per group
```

## Network — `type = "network"`

The directed transition network. The default weight is the **residual**,
so colour means something (warm = over-represented, cool = avoided). The
non-signed weights (`"count"`, `"prob"`, `"lift"`) are all non-negative,
so they draw in a single neutral colour with magnitude carried by edge
width.

```{r network, eval = has_cg}
plot(fit, type = "network")                       # single fit
plot(fit, type = "network", weights = "count")    # frequency network (width)
```

```{r network-group, eval = has_cg}
plot(gf, type = "network")                        # grouped: one per group
```

`prob` (row-conditional) and `count` (raw frequency) are the same graph
with different edge weights: `prob` normalises each source's out-edges to
sum to 1, which promotes rare-source transitions relative to busy ones.

## Chord — `type = "chord"`

States are arcs on a ring; each transition is a filled ribbon (width =
frequency, fill = residual).

```{r chord, eval = has_cg}
plot(fit, type = "chord")                         # single fit
```

```{r chord-group, eval = has_cg}
plot(gf, type = "chord")                          # grouped: one per group
```

## Bootstrap forest — `plot()` on a bootstrap

Resampling has its own circular view: each transition is a spoke spanning
its bootstrap confidence interval, a square marks the estimate, and a
dashed ring marks the null. Stable-residual edges are coloured by
direction; the rest are grey.

```{r forest, eval = has_gg}
b <- bootstrap_lsa(fit, R = 300)
plot(b)                          # residual CIs, circular
plot(b, show_nonsig = FALSE)     # significant transitions only
```

Bootstrap one group at a time to compare them:

```{r forest-group, eval = has_gg}
plot(bootstrap_lsa(gf$A, R = 300))   # group A's bootstrap forest
```

## The worker functions

`plot(fit, type = )` is a thin front door over four exported workers, in
case you want a view's full argument list inline:

```{r workers, eval = FALSE}
plot_transitions(fit, weights = "residuals", significant = TRUE)  # network
plot_chords(fit, width = "prob")                                  # chord
plot_polar(fit, style = "wedge", labels = "all")                  # sunburst
plot_forest(bootstrap_lsa(fit), n_top = 12)                       # forest
```

## In short

```r
plot(fit)                        # residual heatmap (default)
plot(fit, type = "network")      # transition network
plot(fit, type = "chord")        # chord diagram
plot(fit, type = "sunburst")     # polar rose
plot(bootstrap_lsa(fit))         # circular bootstrap forest
plot(gf)                         # grouped: one panel per group (any type)
```
