---
title: "A complete workflow: from sequences to a group comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A complete workflow: from sequences to a group comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 150, fig.width = 7,
                      fig.height = 5.6, out.width = "100%",
                      fig.align = "center")
set.seed(2026)
library(lagseq)
options(digits = 3)
has <- function(p) requireNamespace(p, quietly = TRUE)
```

Lag-sequential analysis (LSA) asks whether, given a state at one moment,
another state follows it more or less often than chance. The method has a
long history in the behavioural sciences, but applied work has often
stopped at description -- reading off observed transition rates -- without
the inferential machinery needed to establish that a pattern reflects the
process rather than the particular sample.

`lagseq` implements LSA as a member of the Dynalytics framework (Saqr,
Lopez-Pernas, and Misiejuk, 2026), whose central tenet is a *scientific
contract*: every analytical claim must be matched by evidence appropriate
to its structure, scope, and complexity, and estimation, validation, and
interpretation are treated as one inferential system rather than separate
steps. Accordingly, in `lagseq` each edge is a *tested departure from
independence* -- the adjusted residual is a test statistic, not a
descriptive weight.

The package is built around a small set of design commitments:

- **Tidy.** Every result is produced by a verb (`transitions()`,
  `nodes()`, `tests()`, `initial()`, ...) that returns a
  one-row-per-observation data frame; the user never indexes into an
  object.
- **Interoperable with the tna family.** `lsa()` ingests sequence-bearing
  objects from `tna`, Nestimate, and `TraMineR` and shares their
  `actor` / `action` / `time` grammar, while a fitted model converts back
  to a `tna` network in one call.
- **Rigorous, following Dynalytics.** The confirmatory testing battery is
  built in: analytic certainty and the bootstrap for edge-level
  uncertainty, split-half reliability for the whole network, case-dropping
  for stability, and permutation under exchangeability for group
  comparison.
- **Expanded.** Beyond the classical engine the package offers several
  model variants, multi-lag analysis, structural-zero constraints, and an
  analytic Bayesian certainty estimator.
- **Visual.** A single `plot()` verb renders the model as a residual
  heatmap, a residual network, a probability-weighted TNA transition
  network, a chord diagram, or a polar sunburst, alongside dedicated plots
  for uncertainty and for group differences.
- **Group-aware.** A grouping column yields one model per group, and
  `compare_lsa()` tests whether the groups differ by permutation.

The remainder of this vignette walks one analysis end to end on real
bundled data, treating each step as one tier of the claim-to-evidence
map: descriptive views support claims about which transitions exist;
certainty intervals and bootstraps support claims about particular edges;
split-half reliability supports claims about the reproducibility of the
whole network; and permutation tests support claims that two groups
differ.

# The data: engagement trajectories

`engagement` is 138 students observed over 15 weekly time points; each
cell is that week's engagement state - `Active`, `Average`, or
`Disengaged`. It is a wide matrix: **one sequence per row**. This section
anchors the scope of every later claim: the units are students, the
events are weekly engagement states, and the possible states are known
before the model is fitted. In Dynalytics terms, this is the descriptive
foundation on which the evidential contract rests. Before asking whether
an edge is surprising or a group differs, we first make clear what
process the model is allowed to represent.

```{r data}
dim(engagement)
head(engagement)
```

# Fit

`lsa()` estimates the model in a single call. It tabulates every
state-to-state transition, derives the frequencies expected under the
independence (no-memory) model, and computes an **adjusted residual** for
each cell: a standardised measure of how far the observed frequency
departs from its expectation. The fitted object is therefore a *residual
network* -- each edge is a tested deviation from independence rather than a
raw rate. This is the named-model discipline of Dynalytics: the same
sequences could be re-expressed as a probability-weighted transition (TNA)
network, but that is a distinct model answering a distinct question -- the
typical next state rather than the surprising one.

```{r fit}
fit <- lsa(engagement)
fit
```

# Read the fit

Every result is produced by a verb that returns a tidy data frame; the
object is never indexed directly. `transitions()` is the primary
accessor, and its arguments select the transitions of interest. This step
supports the most modest claim in the workflow -- which departures from
chance are present in the fitted model. The evidence is descriptive and
edge-level, appropriate for reading the model but not yet for claims about
stability, replication, or group differences.

```{r read}
transitions(fit, significant = TRUE)   # transitions beyond chance
```

A positive residual means the transition happens **more** than chance
(over-represented); a negative one means it is **avoided**. The other
results follow the same one-verb pattern. The companion tables provide
the base rates, node volumes, starting states, and tablewise tests that
keep interpretation grounded in the observed process. They help prevent
an edge from being read in isolation, which is part of the Dynalytics
idea that estimation and interpretation must stay connected.

```{r read2}
nodes(fit)      # per-state incoming / outgoing volume
tests(fit)      # tablewise independence tests
initial(fit)    # where sequences start
```

# Plot - the full gallery

Every view is produced by `plot(fit, type = )`, and each answers a
different descriptive question. Visualisation is the first evidence tier:
it can support claims about pattern, concentration, and direction, but it
is not confirmatory evidence about uncertainty or group differences.

## Residual heatmap

This is the default view. Rows are the current state, columns the next,
and colour is the adjusted residual, so the heatmap displays the
departure-from-chance model directly. Strong colours mark transitions
whose observed counts sit far from the independence expectation, which
makes the view well matched to descriptive claims about where the process
departs from chance.

```{r heatmap, eval = has("ggplot2")}
plot(fit)                                   # type = "heatmap"
```

## Residual network

The same residuals as a directed graph. The convention matches TNA and
Nestimate: **blue = more than chance** (solid edges), **red = less**
(dashed, with a soft halo). It shows *which* transitions are surprising.
The graph keeps the residual scale but makes direction and connectedness
easier to read. It is still the RESIDUAL model: blue and red edges are
not simply popular or unpopular transitions, but tested deviations from
what independence would predict.

```{r net-residual, eval = has("cograph")}
plot(fit, type = "network")
```

## Transition network (a TNA model)

Weighting the edges by probability instead yields the familiar transition
network. `lagseq` constructs a `tna` object on the fly and renders it with
tna's styling: coloured nodes, initial-probability arcs, and weighted
directed edges. The named model changes even though the underlying
sequences do not. A transition (TNA) network is a first-order Markov model
whose edge weights are conditional probabilities; it is the appropriate
display when the claim concerns the typical next state, whereas the
residual network is appropriate when the claim concerns a tested departure
from chance.

```{r net-transition, eval = has("tna")}
plot(fit, type = "network", weights = "prob")
```

## Chord and sunburst

Two further views connect local edges to the overall movement of the
process: a chord diagram of the transition flow, and a polar sunburst of
each state's outgoing distribution. Both are descriptive - for example,
they make clear when one state distributes its outgoing transitions
broadly while another concentrates them in a single destination. They do
not replace the uncertainty checks that follow; they make the fitted
structure easier to inspect before stronger claims are made.

```{r chord, eval = has("cograph")}
plot(fit, type = "chord")
```

```{r sunburst, eval = has("ggplot2")}
plot(fit, type = "sunburst")
```

# Verify and validate

To validate the trustworthiness of an estimated edge, its uncertainty
must be quantified. This is where the Dynalytics contract becomes
explicit: a claim about a specific edge requires edge-level evidence,
because a single observed value is weaker support than a measure of how
much that value could vary. `lagseq` provides two complementary
procedures, both returning a tidy object.

`certainty_lsa()` is the analytic procedure. It places a
Dirichlet-Multinomial posterior on each state's outgoing transitions,
which gives an exact credible interval for every transition probability in
closed form, without resampling. The interval states directly how
precisely the data pin down each edge, and so provides the evidence
appropriate to a claim about a specific transition.

```{r certainty}
certainty_lsa(fit)
as.data.frame(certainty_lsa(fit)) |> head(4)
```

`bootstrap_lsa()` is the resampling counterpart. It resamples whole
sequences with replacement, re-estimates the model on each resample, and
summarises how much each edge varies. The two procedures agree when the
population is homogeneous; the bootstrap is preferred when it is a
mixture, because resampling whole sequences preserves the within-sequence
dependence that the analytic model treats as independent. The forest plot
below shows each edge's interval. Both procedures supply the edge-level
evidence a claim about a single transition requires: a narrow, stable
interval indicates the transition is not an artefact of the particular
sample.

```{r bootstrap}
as.data.frame(bootstrap_lsa(fit, R = 200)) |> head(4)
```

```{r forest, eval = has("ggplot2")}
plot(bootstrap_lsa(fit, R = 200))           # circular forest of edge CIs
```

`reliability_lsa()` raises the claim from a single edge to the whole
network. It repeatedly splits the sequences into two halves, estimates a
model on each half, and correlates the two edge-weight vectors. A high
average correlation indicates that the network is reproducible within the
sample; a low one indicates that the structure is sample-dependent and
should be interpreted with caution. This evidence matches a claim about
the network as a whole rather than about any single edge.

```{r reliability}
reliability_lsa(fit, R = 30)
```

# Starting from a raw event log

`engagement` arrived as ready-made sequences. Real data is often a **long
event log** instead - one row per event. `lsa()` sequences it on the fly
with the same `actor` / `action` / `time` grammar that `tna` uses.
This step shows that the same inferential chain can begin from raw
process traces rather than from an already assembled sequence matrix.
The data format changes, but the contract does not: define the actors,
events, and ordering rule before estimating a model from them.

`tna::group_regulation_long` is exactly that: 2,000 students, each a
stream of timestamped regulation actions, with a recorded achievement
group. The grouping variable will matter later, but it is first treated
as metadata attached to recovered sequences. Keeping sequencing and
grouping explicit helps preserve the units that later resampling and
permutation procedures require.

```{r log, eval = has("tna")}
log <- tna::group_regulation_long
head(log)
```

A single call converts the log into a fitted model: `lsa()` groups events
by actor, orders them by time, splits sessions on long gaps, and fits.
Those operations are not clerical details; they define the empirical
process to which the residual test is applied. Once the log has been
converted to ordered transitions, the fitted edges again represent
departures from independence in that recovered sequence process.

```{r log-fit, eval = has("tna")}
fit_log <- lsa(log, actor = "Actor", action = "Action", time = "Time")
fit_log
```

That grammar is shared deliberately, so the objects interoperate. The
data can be prepared with tna and passed directly to `lsa()`, or a fit can
be converted the other way for tna tooling.
Interoperability is useful because Dynalytics separates the model from
the rendering and analysis layers. The same empirical object can be read
as a residual lag-sequential model when testing departures from chance,
or passed into transition-network tooling when the question is about
probability-weighted flow.

```{r interop, eval = has("tna")}
lsa(tna::tna(tna::prepare_data(log, actor = "Actor",
                               action = "Action", time = "Time")))
lsa_to_tna(fit_log, weights = "prob")       # fit -> a tna network object
```

# Group

To compare achievement groups, name the grouping **column** in the same
call - `group = "Achiever"`. `lsa()` derives one label per recovered
sequence (the column must be fixed within each actor), so there is no
manual splitting or relabelling. Grouped fitting estimates one
lag-sequential model per group under the same data grammar, which makes
the models comparable. At this point the claim is still descriptive:
each group has its own fitted residual structure.

```{r group, eval = has("tna")}
gfit <- lsa(log, actor = "Actor", action = "Action", time = "Time",
            group = "Achiever")
gfit
```

Every verb now returns grouped, tidy results with a leading `group`
column. This makes it easy to inspect apparent differences without
turning visual contrast into inference. Side-by-side networks almost
always differ somewhere, so the next step asks whether those differences
are larger than expected under a valid null.

```{r group-read, eval = has("tna")}
transitions(gfit, significant = TRUE) |> head(4)
```

# Compare the groups

`compare_lsa()` tests whether the two groups have a different transition
structure. It permutes the group labels and compares an N-invariant
effect size (the log-odds ratio), returning a per-edge table together with
a single omnibus test of whether the groups differ at all.
This is the evidence tier appropriate to a comparative claim. Under the
null hypothesis of no group difference, group labels are exchangeable, so
shuffling them builds a reference distribution for any chosen difference
function. The observed edge differences and the omnibus statistic are
then judged against that permutation distribution rather than against
the visual fact that two separately estimated networks are not identical.

```{r compare, eval = has("tna")}
cmp <- compare_lsa(gfit, R = 500, adjust = "BH")
cmp
```

The back-to-back **barrel** displays the result: each group's bar runs to
one side, the higher group's bar is bordered (darker for a larger
difference), and the centre chip carries the difference p-value. The plot
is a compact rendering of the permutation result, not a substitute for it:
the bars show direction and magnitude, while the centre chip records
whether the observed difference is large relative to the differences
produced by label shuffling under exchangeability.

```{r barrel, eval = has("tna") && has("ggplot2")}
plot(cmp)
```

The same comparison as a full-grid difference heatmap. Use it when the
claim is easier to assess as a matrix of source-by-target contrasts than
as paired bars. The underlying evidence remains the same permutation
comparison; only the visual summary changes.

```{r diff-heatmap, eval = has("tna") && has("ggplot2")}
plot(cmp, style = "heatmap")
```

# In short

The compact recipe below is the same workflow expressed as a checklist:
estimate the named model, read and visualise its descriptive structure,
validate edge-level uncertainty, move to logs when needed, preserve group
labels at the sequence level, and use permutation for group comparisons.
Each step escalates the evidence only when the claim being made also
escalates.

```r
fit <- lsa(engagement)                          # 1. fit ready sequences
transitions(fit, significant = TRUE)            # 2. read (one verb per result)
plot(fit)                                       # 3. plot: heatmap / network /
plot(fit, type = "network", weights = "prob")   #    chord / sunburst, TNA network
certainty_lsa(fit); bootstrap_lsa(fit)          # 4. validate (analytic + resampling)

log  <- tna::group_regulation_long              # 5. or start from a raw log
fit  <- lsa(log, actor =, action =, time =)     #    TNA grammar, one call
gfit <- lsa(log, actor =, action =, time =, group = "Achiever")  # 6. group by a column
compare_lsa(gfit)                               # 7. compare + plot(cmp)
```

Taken together, the workflow is a complete inferential chain. Descriptive
claims receive descriptive support, edge claims receive edge-level
uncertainty, whole-network claims receive reproducibility evidence, and
group-difference claims receive permutation evidence under
exchangeability. That is the Dynalytics scientific contract in practice:
every interpretation is tied to the evidence tier that can actually
support it.
